The Origins of Congestion and\\Network Assisted End-to-End Congestion Control

### 2020-05-27 paper/paper.tex
Some more improvements

% author
\author{\IEEEauthorblockN{Til Mohr}
\IEEEauthorblockA{RWTH Aachen University\\til.mohr@rwth-aachen.de}}

@@

\section{Introduction}
Congestion was, is, and always will be a problem on the Internet, appearing in various forms with different symptoms causing minor, but also major challenges. Most data sent over the internet is split up in smaller packets that all try to reach their destination the fastest, whether it is by going the shortest or the least congested route. In an optimal network all computers, servers, etc. would have a direct link to each other, thus reducing the importance of bandwidth and  consequently removing the need for buffers at all. However, the larger the network grows, the number of links would increase quadratically.
\\Networks today consist of network nodes (routers, switches, ...) which all try to route internet packets as quickly as possible whilst keeping the required hardware to a minimum. Ideally there would be a steady flow of packets being transmitted at maximum rate. However, when a internet node receives a packet while the outgoing connection still is in use, the packet has to be stored in a buffer until the outgoing connection freed up again. Therefore, those storage units are essential to every node on a network, but without proper management even well-sized buffers could lead to network problems like packet-loss, jitter, and connection delay \cite{appenzeller2004sizing}.
\\This paper will present origins of network congestion and analyze how different types of algorithms try to avoid and minimize congestion.

\section{Origins of Congestion}
With the ever-growing amount of internet services not only data-centers, but the whole internet in general faces new challenges with each packet sent. Therefore, a network-wide reduction of those complications is an important task to improve the quality of all internet services.
\\Network delay can generally speaking be categorized into three different types of delay \cite{gettys2012bufferbloat}:

@@

\item
Each link creates an intentional gap between two packets being sent. This pause is often necessary to allow the receiver to prepare for a new package, however, depending on the transmission protocol, other purposes may apply. This kind of delay is known as \textit{Inter-Packet delay}.

@@

It is important to understand that delay has a direct impact on congestion. When delay increases, the networks throughput can decrease as packets will stay inside the network for a longer time. If the network however continues to receive new packets at constant rate, more and more packets will have to be stored, resulting in congestion. Generally speaking, to tackle congestion, firstly delay can be reduced, and secondly the networks storages, the buffers, can be improved regarding efficient queue management and buffer size. Finally, transmission protocols can also react to congestion and adapt their send rate to recover from congestion [\nameref{sec:ca}] and even control congestion [\nameref{sec:cc}].
\\Whereas processing delay is almost not noticeable and transmission and inter-packet delay could be improved by upgrading cables or transmission protocols, queuing delay has various causes and therefore is the  most difficult one to reduce. There are many different ways to implement buffers in a network node and the effectiveness of each implementation varies from where it is used. Traffic in residential network nodes for example differs from nodes used in large data-centers in every aspect, thus hardware and software in those components should be adjusted accordingly.
\\Sizing buffers is a complex task. On the one hand read and write times should be as short as possible to minimize additional delay. On the other hand buffers should not be too small, as networks need to be robust for future changes and therefore nodes should store lots of data if needed, while also keeping in mind the requirements of congestion detection algorithms as described in section \nameref{sec:ca} in order to adjust for congestion appropriately \cite{staff2012bufferbloat}. Unifying those goals is very challenging as it leads to increasing costs and physical space occupied by those buffers. Additionally, a 2004 paper by Appenzeller et al. \cite{appenzeller2004sizing} found that 99\% of buffers in network nodes, which were sized on a rule-of-thumb rule based on the TCP protocol \cite{villamizar1994high}, could be removed without losing, but rather gaining performance. Clearly, balancing speed and size is difficult and thus buffer sizes should be adjusted to their area of use accordingly.
\\This isn't the only major cause of queuing delay. Without proper buffer management even nodes with well-sized buffers will experience performance loss. Proper queue management is able to compensate for the unpredictable fluidity of network traffic. It is a difficult task to buffer the right amount of packages to keep transmission flow consistent over time.
As the number of packets being transmitted grows, the \textit{throughput} of the network node increases until packets are being transmitted at bottleneck rate. After this, the transmission rate cannot increase anymore, which leads to packets being buffered along their path to prevent them from being lost \cite{gettys2012bufferbloat}. If the incoming amount of traffic keeps to exceed the amount of outgoing traffic, those buffers will fill up (\textit{over-buffering}), which will eventually lead to \textit{bufferbloat} \cite{Allman13commentson,6886125}.

@@

\section{Congestion Avoidance}
\label{sec:ca}
As already mentioned, congestion detection algorithms make up an important part in congestion avoidance protocols. \textit{TCP} is such a protocol, based on the transport layer, which adapts its send rate according to available network bandwidth \cite{1209197,jacobson1992tcp}. To fully understand, how TCP's congestion avoidance algorithm operates, one needs to understand how TCP works. In the following TCP's congestion avoidance algorithm will be explained using the TCP Specification \cite{ietf-tcpm-rfc793bis-16,postel1981transmission}, also being refered to as \textit{classic TCP}. Note, there are also other TCP implemtations improving upon it. The underlying concept however is the same.
\\TCP lives on the transport layer. It communicates with the application layer via sockets, splits up data into packets (and vice versa), and transmits those packages over the network with the help of the internet protocol, thus TCP is often also referred to as \textit{TCP/IP}. It is the interface between network components and computer programs. Each of those packets have different purposes, but underlay the same structure - the protocol header. The TCP header is split up into multiple segments [Figure \ref{fig:TCP_Header}]:
\begin{figure}
\centering
\begin{bytefield}[bitheight=2.2\baselineskip ,bitwidth=0.635\baselineskip]{32}
	\bitbox{16}{Source Port} & \bitbox{16}{Destination Port} \\
	\bitbox{32}{Sequence Number} \\
	\bitbox{32}{Acknowledgment Number} \\
	\bitbox{4}{Data Offset} & \bitbox{4}{Rsrvd} &
\bitbox{1}{\tiny C\\W\\R} & \bitbox{1}{\tiny E\\C\\E} &
\bitbox{1}{\tiny U\\R\\G} & \bitbox{1}{\tiny A\\C\\K} &
\bitbox{1}{\tiny P\\S\\H} & \bitbox{1}{\tiny R\\S\\T} &
\bitbox{1}{\tiny S\\Y\\N} & \bitbox{1}{\tiny F\\I\\N} &
\bitbox{16}{Window Size} \\
\bitbox{16}{Checksum} & \bitbox{16}{Urgent Pointer} \\
\bitbox{32}{Options} \\
\bitbox{32}{Data}
\end{bytefield}
\caption{Basic structure of the TCP header. It is split up into multiple 32-bit blocks. The data block contains all the application data to be transmitted \cite{jacobson1992tcp,ietf-tcpm-rfc793bis-16,huston2000tcp}.}
\label{fig:TCP_Header}
\end{figure}
\begin{itemize}
\item \textit{Source Port} and \textit{Destination Port} specify the ports on which TCP should communicate with the application layer.

@@

\subsection{Adjusting for Congestion}
\label{sec:AfC}
To prevent congestion altogether, packet flow inside a network would have to be conservative. A connection in this state is also being described as being \textit{in equilibrium} \cite{jacobson1992tcp}. This simply means, the networks works at full load whilst no additional delay or congestion is being created. However, it is quite impossible to keep a full network in this state forever. The goal of TCP's congestion avoidance algorithms, which are based on an \textit{Additive-Increase Multiplicative-Decrease} (AIMD) principle, is to react to the current state of congestion and achieve equilibrium. Those algorithms can be categorized into three sub-algorithms, that all have the goal to keep the connection in equilibrium, no matter how congestion was perceived. Here, a variable called \textit{congestion window} (cwnd) has an important role. It describes the maximum amount of data sent during an average RTT. An additional mechanism of keeping a connection in equilibrium is, that each acknowledgment packet received by the sender opens the space for a new packet to be sent into the network \cite{jacobson1992tcp}. The principles TCP uses to adjust for congestion are based on \textit{Additive-Increase Multiplicative-Decrease} (AIMD) principle.

@@

\subsection{Evaluation of Network Assisted Congestion Control}
As you can see, including the network layer to detect congestion can have great performance improvements. 'Understanding XCP: Equilibrium and Fairness' has shown through simulations that many network topologies could benefit from XCP as it clears queues in equilibrium, thus resulting in improved network performance under load \cite{1498331}. Another paper has shown that XCP outperforms ordinary TCP in nearly any environment related to average queue lengths, efficiency, and throughput \cite{katabi2002congestion}.
\\However, all this performance comes at a cost. Since network nodes have to actively detect congestion, ECN, XCP, and other network assisted congestion control protocols rely on active queue management in those nodes. Although there is some hardware being produced today that meet ECNs requirements, a large part of internet hardware isn't capable of doing so. As many routers do not run on software but hardware only to improve performance, they can only handle specific packets. Whenever they cannot recognize a packet, it will simply be dropped \cite{katabi2002congestion}. So, although ECN packets differ from ordinary TCP/IP packets by only one bit, not all routers can handle this change and will drop those packets. This is the reason you do not see those protocols in wide area networks (WAN) too often, let alone the internet.
\\Nevertheless, whenever the implementation of XCP, ECN, or similar protocols is not a challenge, you can implement network assisted congestion control with ease and see great performance improvements. Especially smaller networks with high traffic could definitely benefit from those technologies [\nameref{sec:DCTCP}].

@@

\subsection{DCTCP}
\label{sec:DCTCP}
A lot of internet services run in data centers with large amounts of computers communicating with each other, resulting in a lot of traffic inside the data center. Optimizing those data centers networking capabilities can result in an increase in application productivity. In 2010 a paper proposed \textit{DCTCP}, a variation of TCP using ECN to create a data transportation protocol, which especially takes hardware purposed for use in those environments into account \cite{10.1145/1851275.1851192}.\\
The general problem that appears in high-traffic environments is the so-called \textit{TCP Incast} problem. It describes a drastic reduction in application throughput. To understand where TCP incast originates, you fist need to understand how most data centers are built. A typical data center contains a set of racks, each holding a large number of servers, the switches, that connect those racks, and connecting links, that connect those switches to other parts of the data center or the internet [Figure \ref{fig:dc}]. This design pattern is also known as \textit{Partition/Aggregate} and is most commonly used in data centers. It however results in a high 'fan-in', as multiple servers (leaves) are connected to just one link (root) \cite{10.1145/1851275.1851192,10.1145/1592681.1592693}. In addition, most applications require high bandwidth, low latency connections, also enabling them to send many parallel requests to other servers. Combining those requirements with the design pattern, it becomes clear, that those switches are being challenged. Besides, many data center switches have relatively small, shared buffers. Therefore, it is common that queue lengths stay high for a prolonged period of time, resulting in a collapse of throughput \cite{10.1145/1851275.1851192,10.1145/1592681.1592693}.\\
The goal of DCTCP is to keep data center traffic constantly flowing, which means meeting the requirements applications take while keeping relatively small buffer sizes inside switches in mind. To achieve this, DCTCP implements ECN. Unlike TCP however, if a DCTCP sender registers congestion, the window size will be reduced by a fraction of the amount of marked packets it received. Therefore, excessive link underutilization can be prevented. In addition, the DCTCP receiver will only send one ACK packet for every $m$ consecutive data packet, where all $1,...,m$ packets before had the same CE-flag configuration. Here, $m$ is a variable initialized while the DCTCP connection is being established and cannot be changed without renegotiation. Therefore, fewer packets are sent through the network, but the TCP sender is still able to identify lost packets and register congestion \cite{10.1145/1851275.1851192}.
\\Being bandwidth greedy transmission protocol, data centers using DCTCP can see a performance drop in nodes that use other transmission protocols such as TCP or ECN. As DCTCP tries to utilize the links as much as possible, non-greedy protocols will be neglected and will see a drop in performance. Thus, data centers should avoid using DCTCP in addition to other protocols to fully make use of DCTCP's capabilities.
\\To sum up, being able to react to congestion immediately enables DCTCP to reduce buffer overflows and timeouts, thus largely resolving incast problems. The paper proposing DCTCP has found, that their protocol increases traffic by 110\% while using 90\% less buffer space compared to TCP \cite{10.1145/1851275.1851192}.

>>>>

% author
%\author{\IEEEauthorblockN{Til Mohr}
%\IEEEauthorblockA{RWTH Aachen University\\til.mohr@rwth-aachen.de}}

@@

\section{Introduction}
Congestion was, is, and always will be a problem on the Internet, appearing in various forms with different symptoms causing minor, but also major challenges. Most data sent over the internet is split up in smaller packets that all try to reach their destination the fastest, whether it is by taking the shortest or the least congested route. In an optimal network all computers, servers, etc. would have a direct link to each other, thus reducing the importance of bandwidth and  consequently removing the need for buffers at all. However, the larger the network grows, the number of links would increase quadratically.
\\Networks today consist of network nodes (routers, switches, ...) which all try to route internet packets as quickly as possible whilst keeping the required hardware to a minimum. Ideally there would be a steady flow of packets being transmitted at maximum rate. However, when a internet node receives a packet while the outgoing connection still is in use, the packet has to be stored in a buffer until the outgoing connection freed up again. Therefore, those storage units are essential to every node on a network, but without proper management even well-sized buffers could lead to network problems like packet-loss, jitter, and connection delay \cite{appenzeller2004sizing}.
\\This paper will present origins of network congestion and analyze how different types of algorithms try to avoid and minimize congestion.

\section{Origins of Congestion}
With the ever-growing amount of internet services not only data-centers, but the whole internet in general faces new challenges with each packet sent. Therefore, a network-wide reduction of those complications is an important task to improve the quality of all internet services.
\\Network delay can generally speaking be categorized into four different types of delay \cite{gettys2012bufferbloat}:

@@

\item
Each node creates an intentional gap between two packets being sent. This pause is often necessary to allow the receiver to prepare for a new package, however, depending on the transmission protocol, other purposes may apply. This kind of delay is known as \textit{Inter-Packet delay}.

@@

It is important to understand that delay has a direct impact on congestion. When delay increases, the networks throughput can decrease as packets will stay inside the network for a longer time. If the network however continues to receive new packets at constant rate, more and more packets will have to be stored, resulting in congestion. Generally speaking, to tackle congestion, firstly delay can be reduced, and secondly the networks storages, the buffers, can be improved regarding efficient queue management and buffer size. Finally, transmission protocols can also react to congestion and adapt their send and data rate to recover from congestion [\nameref{sec:ca}] and even control congestion [\nameref{sec:cc}].
\\Whereas processing delay is almost not noticeable and transmission and inter-packet delay could be improved by upgrading cables, hardware, or transmission protocols, queuing delay has various causes and therefore is the  most difficult one to reduce. There are many different ways to implement buffers in a network node and the effectiveness of each implementation varies from where it is used. Traffic in residential network nodes for example differs from nodes used in large data-centers in every aspect, thus hardware and software in those components should be adjusted accordingly.
\\Sizing buffers is a complex task. On the one hand read and write times should be as short as possible to minimize additional delay. On the other hand buffers should not be too small, as networks need to be robust for future changes and therefore nodes should store lots of data if needed, while also keeping in mind the requirements of congestion detection algorithms as described in section \nameref{sec:ca} in order to adjust for congestion appropriately \cite{staff2012bufferbloat}. Unifying those goals is very challenging as it leads to increasing costs and physical space occupied by those buffers. Additionally, a 2004 paper by Appenzeller et al. \cite{appenzeller2004sizing} found that 99\% of buffers in network nodes, which were sized on a rule-of-thumb rule based on the TCP protocol \cite{villamizar1994high}, could be removed without losing, but rather gaining performance. Clearly, balancing speed and size is difficult and thus buffer sizes should be adjusted to their area of use accordingly.
\\This isn't the only major cause of queuing delay. Without proper buffer management even nodes with well-sized buffers will experience performance loss. Proper queue management is able to compensate for the unpredictable fluidity of network traffic. It is a difficult task to buffer the right amount of packages to keep transmission flow consistent over time.
As the number of packets being transmitted grows, the \textit{throughput} of the network node increases until packets are being transmitted at bottleneck rate. After this, the transmission rate cannot increase anymore, which leads to packets being buffered along their path to prevent them from being lost \cite{gettys2012bufferbloat}. If the incoming amount of traffic keeps to exceed the amount of outgoing traffic, those buffers will fill up (\textit{over-buffering}), which will eventually lead to \textit{bufferbloat} \cite{Allman13commentson,6886125}.

@@

\section{Congestion Avoidance}
\label{sec:ca}
As already mentioned, congestion detection algorithms make up an important part in congestion avoidance protocols. \textit{TCP} is such a protocol, based on the transport layer, which adapts its send rate according to available network bandwidth \cite{1209197,jacobson1992tcp}. To fully understand, how TCP's congestion avoidance algorithm operates, one needs to understand how TCP works. In the following, TCP's congestion avoidance algorithm will be explained using the TCP Specification \cite{ietf-tcpm-rfc793bis-16,postel1981transmission}, also being refered to as \textit{classic TCP}. Note, there are also other TCP implemtations improving upon it. The underlying concept however is the same.
\\TCP lives on the transport layer. It communicates with the application layer via sockets, splits up data into packets (and vice versa), and transmits those packages over the network with the help of the internet protocol, thus TCP is often also referred to as \textit{TCP/IP}. It is the interface between network components and computer programs. Each of those packets have different purposes, but underlay the same structure - the protocol header. The TCP header is split up into multiple segments [Figure \ref{fig:TCP_Header}]:
\begin{figure}
\centering
\begin{bytefield}[bitheight=2.2\baselineskip ,bitwidth=0.635\baselineskip]{32}
	\bitbox{16}{Source Port} & \bitbox{16}{Destination Port} \\
	\bitbox{32}{Sequence Number} \\
	\bitbox{32}{Acknowledgment Number} \\
	\bitbox{4}{Data Offset} & \bitbox{4}{Rsrvd} &
\bitbox{1}{\tiny C\\W\\R} & \bitbox{1}{\tiny E\\C\\E} &
\bitbox{1}{\tiny U\\R\\G} & \bitbox{1}{\tiny A\\C\\K} &
\bitbox{1}{\tiny P\\S\\H} & \bitbox{1}{\tiny R\\S\\T} &
\bitbox{1}{\tiny S\\Y\\N} & \bitbox{1}{\tiny F\\I\\N} &
\bitbox{16}{Window Size} \\
\bitbox{16}{Checksum} & \bitbox{16}{Urgent Pointer} \\
\bitbox{32}{Options} \\
\bitbox{32}{Data}
\end{bytefield}
\caption{Basic structure of the TCP header. It is split up into multiple 32-bit blocks. The data block contains all the application data to be transmitted \cite{jacobson1992tcp,ietf-tcpm-rfc793bis-16,huston2000tcp}.}
\label{fig:TCP_Header}
\end{figure}
\begin{itemize}
\item \textit{Source Port} and \textit{Destination Port} specify the ports on which TCP should communicate with the application layer.

@@

\subsection{Adjusting for Congestion}
\label{sec:AfC}
To prevent congestion altogether, packet flow inside a network would have to be conservative. A connection in this state is also being described as being \textit{in equilibrium} \cite{jacobson1992tcp}. Simply speaking this means, the network works at full load whilst no additional delay or congestion is being created. However, it is quite challenging to keep a full network in this state forever. The goal of TCP's congestion avoidance algorithms, which are based on an \textit{Additive-Increase Multiplicative-Decrease} (AIMD) principle, is to react to the current state of congestion and achieve equilibrium. Those algorithms can be categorized into three sub-algorithms, that all have the goal to keep the connection in equilibrium, no matter how congestion was perceived. Here, a variable called \textit{congestion window} (cwnd) has an important role. It describes the maximum amount of data sent during an average RTT. An additional mechanism of keeping a connection in equilibrium is, that each acknowledgment packet received by the sender opens the space for a new packet to be sent into the network \cite{jacobson1992tcp}. The principles TCP uses to adjust for congestion are based on \textit{Additive-Increase Multiplicative-Decrease} (AIMD) principle.

@@

\subsection{Evaluation of Network Assisted Congestion Control}
As you can see, including the network layer to detect congestion can have great performance improvements. 'Understanding XCP: Equilibrium and Fairness' has shown through simulations that many network topologies could benefit from XCP as it clears queues in equilibrium, thus resulting in improved network performance under load \cite{1498331}. Another paper has shown that XCP outperforms ordinary TCP in nearly any environment in aspects of average queue lengths, efficiency, and throughput \cite{katabi2002congestion}.
\\However, all this performance comes at a cost. Since network nodes have to actively detect congestion, ECN, XCP, and other network assisted congestion control protocols rely on active queue management in those nodes. Although there is some hardware being produced today that meet ECNs requirements, a large part of internet hardware isn't capable of doing so. As many routers do not run on software but hardware only to improve performance, they can only handle specific packets. Whenever they cannot recognize a packet, it will simply be dropped \cite{katabi2002congestion}. So, although ECN packets differ from ordinary TCP/IP packets by only one bit, not all routers can handle this change and will drop those packets. This is the reason you do not see these kind of protocols in wide area networks (WAN) too often, let alone the internet.
\\Nevertheless, whenever the implementation of XCP, ECN, or similar protocols is not a challenge, you can implement network assisted congestion control with ease and see great performance improvements. Especially smaller networks with high traffic could definitely benefit from those technologies [\nameref{sec:DCTCP}].

@@

\subsection{DCTCP}
\label{sec:DCTCP}
A lot of internet services run in data centers with large amounts of computers communicating with each other, resulting in a lot of traffic inside the data center. Optimizing those data centers networking capabilities can result in an increase in application productivity. In 2010 a paper proposed \textit{DCTCP}, a variation of TCP using ECN to create a data transportation protocol, which especially takes hardware purposed for use in those environments into account \cite{10.1145/1851275.1851192}.\\
The general problem that appears in high-traffic environments is the so-called \textit{TCP Incast} problem. It describes a drastic reduction in application throughput. To understand where TCP incast originates, you fist need to understand how most data centers are built. A typical data center contains a set of racks, each holding a large number of servers, the switches, that connect those racks, and connecting links, that connect those switches to other parts of the data center or the internet [Figure \ref{fig:dc}]. This design pattern is also known as \textit{Partition/Aggregate} and is most commonly used in data centers. It however results in a high 'fan-in', as multiple servers (leaves) are connected to just one link (root) \cite{10.1145/1851275.1851192,10.1145/1592681.1592693}. In addition, most applications require high bandwidth, low latency connections, also enabling them to send many parallel requests to other servers. Combining those requirements with the design pattern, it becomes clear, that those switches are being challenged. Besides, many data center switches have relatively small, shared buffers. Therefore, it is common that queue lengths stay high for a prolonged period of time, resulting in a collapse of throughput \cite{10.1145/1851275.1851192,10.1145/1592681.1592693}.\\
The goal of DCTCP is to keep data center traffic constantly flowing, which means meeting the requirements applications take while keeping relatively small buffer sizes inside switches in mind. To achieve this, DCTCP implements ECN. Unlike TCP however, if a DCTCP sender registers congestion, the window size will be reduced by a fraction of the amount of marked packets it received. Therefore, excessive link underutilization can be prevented. In addition, the DCTCP receiver will only send one ACK packet for every $m$ consecutive data packet, where all $1,...,m$ packets before had the same CE-flag configuration. Here, $m$ is a variable initialized while the DCTCP connection is being established and cannot be changed without renegotiation. Therefore, fewer packets are sent through the network, but the TCP sender is still able to identify lost packets and register congestion \cite{10.1145/1851275.1851192}.
\\However, being bandwidth greedy transmission protocol, data centers using DCTCP can see a performance drop when they additionally use other transmission protocols such as TCP or ECN. As DCTCP tries to utilize the links as much as possible, non-greedy protocols will be neglected and will see a drop in performance. Thus, data centers should avoid using DCTCP in addition to other protocols to fully make use of DCTCP's capabilities.
\\To sum up, being able to react to congestion immediately enables DCTCP to reduce buffer overflows and timeouts, thus largely resolving incast problems. The paper proposing DCTCP has found, that their protocol increases traffic by 110\% while using 90\% less buffer space compared to TCP \cite{10.1145/1851275.1851192}.
